Formal Paper Draft

### 2021-08-22 PaperDraft.Rmd
mas changeados

---
title: "Formal Paper Draft"
subtitle: "Influence Diagnostics for High-Dimensional Ordination Techniques"
author: "Paul Harmon"
date: "8/21/2021"
output: html_document
---

@@

# High-dimensional Data Visualization with t-SNE

@@

## Problems with using t-SNE

Although t-SNE is highly effective at creating 2 (or 3) dimensional ordinations of high-dimensional data, the mathematical

Finally, performing t-SNE on large datasets can be computationally slow, necessitating a random-walk version that is discussed in a later paper by van der maaten (2014).

## Reduction of Initial Dimensionality with PCA

@@

# Other Multivariate Dimension-Reduction Tools
## MDS
## Sammon Mapping
## UMAP

@@

In regression models, the concept of influential points is

## Influence in

# Methods

This document assesses the effect of multivariate outliers on the maps produced by two methods: classical multidimensional scaling and t-SNE. In general, we seek to identify either **multivariate outliers** (or groups of them) or alternatively single influential points that, for some reason, cause large differences in the shape of the resulting ordination.

Our Permanova-based approach is as follows:

1. For each observation in the dataset, remove it, and generate a t-SNE map in 2 dimensions from the resulting (n-1) observations.  (For accounting purposes, it may make sense to keep the nx1 structure but replace the holdout observation with NA). 
2. Generate a distance matrix of the nx1 by 2 matrix. Do this for each of the held-out observations. 
3. Generate a pairwise-complete correlation matrix from the combination of these vectorized distances that measures similarity of distances (where available).
4. Convert that correlation matrix (of all the hold-out vectorized distance matrices) into a distance matrix using a $1-|\sqrt{}\. 
5. Apply non-parametric PERMANOVA tests for each holdout observation - this means that on a 100-observation dataset, you apply 100 individual tests where each test uses an independent variable that is equal to 1 for the index of the holdout observation and 0 elsewhere. The hypotheses are specified as such:

@@

# Results

>>>>

---
title: "Formal Paper Draft"
author: "Paul Harmon"
date: "8/21/2021"
output:
  pdf_document: default
  html_document: default
subtitle: Influence Diagnostics for High-Dimensional Ordination Techniques
---

@@

# High-dimensional Data Visualization Tools

## t-SNE

@@

### Reduction of Initial Dimensionality with PCA

@@

## Multidimensional Scaling

Classical multidimensional scaling (MDS) is closely related to PCA (an in some cases, equivalent) (Gower, 1966).  MDS takes as input a distance or dissimilarity matrix representing the distance between pairs of observations - this distance need not simply be Euclidean. Classical multidimensional scaling returns maps in teh same metric as the original distance or dissimilarity matrix.

@@

In regression models, an outlier is typically denoted as a point that is either much larger or smaller than similar observations on a specific scale. However, an outlier need not be an influential point. An influential point simply needs to be outlying enough relative to similar observations that it impacts the fit of the regression line. the identification of influential points that can impact the results of a model is commonplace.

Cook (1977) proposed a measure of influence called "Cook's Distance" that has since become one of the most common tools for diagnosing influential points and outliers. The idea is predicated on the concept that by leaving an out an influential observation, the results of the model should change. This process of jacknifing observations is done for all the observations in the data.

Cook's distance is defined as follows for the $ith$ observation, with$\hat{j}_{j}$ the fitted response value for the model withthe $ith$ observation omitted, $s^2$ defined as the mean-squared error for the model and $p$ the number of variables for each observation: 
$$ D_i = \frac{\sum_{j=1}^{n} (\hat{y_j} - \hat{y}_{j(i)})^2}{ps^2}$$

Although Cook's distance is essentially an F statistic from an $F_{p, n-p}$ distribution, and the method does not use p-values to determine how "influential" a point is. Rather, cutoff values are typically assessed, or observations with relatively large Cook's distance values are considered for investigation of influence.

```{r}
#image of an influential point
x = rnorm(100, 10, 3)
y = 3 + 2*x + rnorm(100, 0, 1)

lm1 = lm(y~x)
par(mfrow = c(2,2))
plot(lm1)

```

## Outliers and Influential Points in Maps

Similar to a regression setting, the resulting maps produced by ordination methods like MDS, t-SNE and the like can be impacted by the inclusion or exclusion of points. In short, an "influential" point in a lower dimensional mapping can be defined as one that, when excluded, drives meaningful differences in the ordination method that lead to a different shaped map as compared to when that point is included. An outlier, on the other hand, might look different in the high-dimensional space (or on a subset of features), but would not meaningfully change the shape of the map when left out.

Joliffe (2002) overviews tools for identifying influential observations in Pricipal Component Analysis, which is a related dimension-reducing technique to MDS. Influence functions (Critchley, 1985; Hampel, 1974; Huber, 1981) can be defined for regression methods as well as for loadings in PCA. However, the literature on more recently-developed methods lacks a similar metric, most notably one that can be used or compared across different mapping methods.

Much of the literature around robustness of mapping results and multivariate outliers centers on dealing with the problem by modifying the dimension reduction technique rather than identifying it. Blouvshtein and Cohen-Or (2018) analyzed the effect of outliers on the maps produced by classical multidimensional scaling methods.  In general, these methods are not particularly robust to the effects of multivariate outlier.  They propose a method to detect outliers prior to the generation of the MDS map so that they can be removed from the data. Their method involves treating the distances input to the algorithm as edges of a complete graph that connects each data point.

These edges $d1, d2, d3$ can then be used to form triangles - in general, outliers tend to *break* the triangle inequality and inliers tend to satisfy it. They recommend a rule to examine the histogram H of $b$, the number of edges of broken triangles, and generate a threshold $\phi$ that is intended to identify inliers vs. outliers.

Two other papers, including Forrero and Giannakis (2012) as well as Kong et al. (2019) frame the problem in a slightly different way. They consider a penalized *stress* metric for optimization, utilizing regularization to better identify outlier points.  In both cases, they are able to identify outliers using a majorization-minimization procedure that iteratively produce more outlier-robust maps of the lower-dimensional embedding of interest.

# Method Overview

We instead propose a method for identifying influential observations that is agnostic to the mapping method.  In addition, our method follows a philosophical framework similar in flavor to Cook's distance for mapping methods; however, it relies on a different set of tools than its regression analogue.

Similar to Cook's Distance, we propose calculating hold-out maps, with each of the 1...n observations held out singly.  Rather than basing the test on a standard F distribution, as is Cook's distance, our method utilizes permutation-based MANOVA (Permanova) tests to calculate pseudo-F test statistics.  Our method borrows from some of the machinery of the mantel test (Mantel, 1967), which is a method for comparing correlation between two matrices.

The proposed approach is as follows:

1. For each observation in the dataset, remove it, and generate a t-SNE map in 2 dimensions from the resulting (n-1) observations.  (For accounting purposes, it may make sense to keep the nx1 structure but replace the holdout observation with NA). 
2. Generate a distance matrix of the nx1 by 2 matrix. Do this for each of the held-out observations. 
3. Generate a pairwise-complete correlation matrix from the combination of these vectorized distances that measures similarity of distances (where available).
4. Convert that correlation matrix (of all the hold-out vectorized distance matrices) into a distance matrix using a $1-\|\sqrt{stuff}$. 
5. Apply non-parametric PERMANOVA tests for each holdout observation - this means that on a 100-observation dataset, you apply 100 individual tests where each test uses an independent variable that is equal to 1 for the index of the holdout observation and 0 elsewhere. The hypotheses are specified as such:

@@

# Simulation

To test the method

# Overview of Real Data Results

## Penguin Data

The Palmer Penguin dataset (Horst et al., 2020) contains information pertaining to three different species of penguins.  The dataset is intended to be an alternative to the ever-popular iris dataset (cite) as each of the penguins form relatively distinct clusters when measured on a series of biometric markers.

## Carnegie Classification

The Carnegie Classification (CC) is a method for classifying institutions of higher education based on several metrics of institutional productivity.  The 2015 version of CC was based on data ranging from doctorates awarded in STEM, Non-STEM, Humanities, Social Sciences, and Other (professional) fields, as well as research expenditures in STEM and Non-STEM fields and two proxies for institutional size (tenurable faculty headcount and research faculty headcount).

The Carnegie Classification is typically based on a scaled version of the data due to the presence of outlier institutions on several metrics.  In particular, some institutions spend a great deal of expenditures on research compared to others, so the actual CC is based on two indices generated using Principal Component Analysis on the ranked data (Cite Harmon, Kosar).

For this analysis, we instead look at scaled versions of the unranked data in part to assess whether or not some of the relative extreme points appear as influential relative to other universities in the group.
